Wavelet Multiresolution Spatio-Temporal Modelling Using the Integro-Difference Equation

### 2011-01-13 Multiresolution_framework/ltx/manuscript.tex
minor changes

% correct bad hyphenation here
% \hyphenation{op-tical net-works semi-conduc-tor}

@@

\author{Parham Aram*,
        Michael Dewar and Visakan Kadirkamanathan,~\IEEEmembership{Member,~IEEE}  % <-this % stops a space
\thanks{P. Aram* and V. Kadirkamanathan are with the Department of Automatic Control and Systems Engineering, University of Sheffield, Sheffield, S1 3JD, U.K. (e-mail:p.aram@sheffield.ac.uk; visakan@sheffield.ac.uk).}% <-this % stops a space
\thanks{M. Dewar is with the School of Informatics, University of Edinburgh, Edinburgh, EH8 9AB, U.K. (e-mail:mike.dewar@inf.ed.ac.uk)}}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

@@

A state-space representation of the IDE model allows the use of the Expectation Maximisation (EM) algorithm to infer both the kernel and the spatial field from the noise corrupted spatio-temporal data. In this context, the EM algorithm is a two part iterative algorithm that yields the maximum likelihood estimate of the spatial mixing kernel parameters and the posterior distribution of the field weights. In the M-step a lower bound $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$ of the complete-data log-likelihood  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$, is maximised under the assumption of the system parameters being $\boldsymbol\theta'$ \cite{McLachlan1997}. In the E-step, given an estimate of $\boldsymbol\theta$, filtering methods can be used to determine the posterior state sequence. Here we compare three techniques namely based on  the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005}, Ensemble Kalman Smoother (EnKS) \cite{Evensen2003} and a heuristic separable RTS to solve the joint estimation problem. For the first method we modify the framework proposed in \cite{Dewar2009} by computing the required elements once instead of at each iteration, the modified algorithm is computationally less demanding. see Appendix ...

@@

\begin{figure}[!h] 
\includegraphics[width=0.5\textwidth]{./Graph/aram1.eps}
\caption{Multiresolution decomposition of a complex spatial mixing kernel (dark line) and the corresponding scaling and wavelet basis functions (light lines).}
\label{fig:SpatialMixingKernel}
\end{figure} 
\begin{figure}[!h] 
\includegraphics[width=0.5\textwidth]{./Graph/aram2.eps}
\caption{Spatial mixing kernel; estimated and true kernels are shown by solid and dashed line respectively; top panel is estimation with maximmum scale $j=0$ and bottom panel is estimation with maximum scale $j=1$. }
\label{fig:SpatialMixingKernelEstimate}
\end{figure}
\begin{figure*}[!hb] 
% \renewcommand*{\thesubfigure}{}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram35.eps}}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram310.eps}}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram350.eps}}\\ 
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram3100.eps}}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram3200.eps}} 
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram3300.eps}}
\caption{Spatial field at a number of time instants; estimated and true fields are shown by solid and dashed line respectively; the top panel of each plot is the estimate with maximmum scale $j=0$ and the bottom panel is the estimate with maximum scale $j=1$.}
\label{fig:1DFieldEstimation}
\end{figure*}

@@

\begin{figure}[!h] 
\includegraphics[width=0.5\textwidth]{./Graph/aram4.eps}
\caption{Variance of the field at each spatial location for $j=0$ (dotted line) and $j=1$ (solid line). The observation locations are described with a cross.}
\label{fig:FieldVariance}
\end{figure} 

@@

\begin{figure}[!t] 
\includegraphics[width=0.5\textwidth]{./Graph/aram5.eps}
\caption{Distribution of the parameter estimate for 2-D Example over 500 runs of the EM algorithm. The dotted line indicates the true parameter.}
\label{fig:2DHist}
\end{figure}
\begin{figure*}[!th] 
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram6a.eps}}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram6b.eps}}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram6c.eps}}\\ 
\caption{Spatial field at t=5; a: True field; b: Estimated field with maximmum scale $j=0$; c: Estimated field with maximmum scale $j=1$.}
\label{fig:2DFieldEstimation1}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram7a.eps}}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram7b.eps}} 
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram7c.eps}}\\
\caption{Spatial field at t=50; a: True field; b: Estimated field with maximmum scale $j=0$; c: Estimated field with maximmum scale $j=1$.}
\label{fig:2DFieldEstimation2}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram8a.eps}}
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram8b.eps}} 
\subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram8c.eps}}
\caption{Spatial field at t=150; a: True field; b: Estimated field with maximmum scale $j=0$; c: Estimated field with maximmum scale $j=1$.}
\label{fig:2DFieldEstimation3}
\end{figure*}

@@

\section*{Reducing the computational complexity}

@@

@@

>>>>

% correct bad hyphenation here
% \hyphenation{op-tical net-works semi-conduc-tor}
% Algorithms
\usepackage{float}
\floatstyle{ruled}
\newfloat{algorithm}{htbp}{loa}%[chapter]
\floatname{algorithm}{Algorithm}

@@

\author{In no particular order: P. Aram, M. Dewar, D.R. Freestone, K.Scerri, D.B. Grayden and V. Kadirkamanathan % <-this % stops a space
\thanks{P. Aram* and V. Kadirkamanathan are with the Department of Automatic Control and Systems Engineering, University of Sheffield, Sheffield, S1 3JD, U.K. (e-mail:p.aram@sheffield.ac.uk; visakan@sheffield.ac.uk).}% <-this % stops a space
\thanks{M. Dewar is with the School of Informatics, University of Edinburgh, Edinburgh, EH8 9AB, U.K. (e-mail:mike.dewar@inf.ed.ac.uk)}}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

@@

A state-space representation of the IDE model allows the use of the Expectation Maximisation (EM) algorithm to infer both the kernel and the spatial field from the noise corrupted spatio-temporal data. In this context, the EM algorithm is a two part iterative algorithm that yields the maximum likelihood estimate of the spatial mixing kernel parameters and the posterior distribution of the field weights. In the M-step a lower bound $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$ of the complete-data log-likelihood  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$, is maximised under the assumption of the system parameters being $\boldsymbol\theta'$ \cite{McLachlan1997}. In the E-step, given an estimate of $\boldsymbol\theta$, filtering methods can be used to determine the posterior state sequence. Here we compare three techniques namely based on  the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005}, Ensemble Kalman smoother (EnKs) \cite{Evensen2003} and a heuristic separable RTS smoother for state estimation problem. For the first method we modify the framework proposed in \cite{Dewar2009} by computing and storing the required elements prior to the iterative estimation algorithm, the modified algorithm is computationally less demanding (see Appendix \ref{App:ReducedRTS}). In Ensemble Kalman filter (EKf) an ensemble of model states, $\mathbf X_t \in \mathbb R^{n_x \times N} \sim \mathcal N(\mathbf x_t,\mathbf P_t)$ is propogated through the state equation to calculate the forecast estimates $\hat{\mathbf x}^f_t$ and the error covariance of the predicted model state, $\mathbf P_t^f$ where $N$ is the number of ensemble members. Then  the model state is updated to compute analysed estimates, $\hat {\mathbf x}_t^a$.  This is important for model with high dimensional state space where standard Kalman filter  is computationally infeasible to implement as the state covariance matrix needs to be stored and propogated in time. The EnKs updates the filtered ensembles at the current and all prior times every time a new set of measurements becomes available. The algorithm is given in Algorithm \ref{alg:EnKS} for completeness. It is essential that the observations be also treated as an ensemble of random variables, $\mathbf Y_t$, having a distribution with mean equal to the observed value, $\mathbf y_t$ and covariance equal to $\boldsymbol\Sigma_v$ \cite{Houtekamer2005}.

% represents the predicted state covariance matrix, $\mathbf P_{t|t+1}$ using a large stochastic ensemble of the predicted states $\hat{\mathbf x}_{t+1|t}$.

\begin{algorithm}
\caption{Ensemble Kalman smoother}
\label{alg:EnKS}
% \begin{tabbing}
\begin{enumerate}
\item Initialise:
\begin{center}
 $\hat{\mathbf x}_0 $, $\mathbf P_0$  
\end{center}
\item
for $t\in\left\lbrace 0, \dots, T\right\rbrace $, calculate the state ensembles $\hat{\mathbf X}_t$ and the model errors ensemble $\mathbf W_t$ and propogate through the state equation:
\begin{equation*}
\hat{\mathbf X}_{i,t+1}^f=\mathbf A \hat{\mathbf X}_{i,t}^a+\mathbf W_{i,t}
\end{equation*}
calculate the predicted state and the predicted covariance matrix
\begin{equation*}
 \hat{\mathbf x}_{t+1}^f=\frac{1}{N}\sum_{i=1}^{N}\hat{\mathbf X}_{i,t+1}^f
\end{equation*}
\begin{equation*}
 \mathbf P_{t+1}^f=\frac{1}{N-1}\sum_{i=1}^{N}\left( \hat{\mathbf X}_{i,t+1}^f- \hat{\mathbf x}_{t+1}^f\right)\left( \hat{\mathbf X}_{i,t+1}^f- \hat{\mathbf x}_{t+1}^f\right)^\top
\end{equation*}
\item
compute $\mathbf X_4(t+1)$ and $\mathbf X_5(t+1)$ and the filtered state and the filtered covariance matrix:
\begin{equation*}
 \mathcal K_t=\mathbf C^{\top}(\mathbf C\mathbf{P}_{t+1}^f\mathbf C^{\top}+\boldsymbol\Sigma_v)^{-1}
\end{equation*}

\begin{equation*}
 \mathbf X_4(t+1)=\frac{1}{N-1}\sum_{i=1}^{N}(\hat{\mathbf X}_{i,t+1}^f-\hat{\mathbf x}^f_{t+1} )^\top\mathcal K_t (\mathbf Y_{i,t+1}-\mathbf C\hat{\mathbf{X}}_{i,t+1}^f)
\end{equation*}
\begin{equation*}
 \mathbf X_5(t+1)= \mathbf I +\mathbf X_4(t+1)
\end{equation*}
\begin{equation*}
 \hat{\mathbf X}_{t+1}^a=\hat{\mathbf X}_{t+1}^f \mathbf X_5(t+1)
\end{equation*}
\item compute the smooth ensemble for $t\in\left\lbrace 0, \dots, T\right\rbrace $
\begin{equation*}
 \hat{\mathbf{X}}_{t}^s=\hat{\mathbf X}_{t}^a \prod_{j=t+1}^{T}X_5(j)
\end{equation*}
\end{enumerate}
% \end{tabbing}
\end{algorithm}

@@

% \begin{figure}[!h] 
% \centering
% \includegraphics[width=0.5\textwidth]{./Graph/aram1.eps}
% \caption{Multiresolution decomposition of a complex spatial mixing kernel (dark line) and the corresponding scaling and wavelet basis functions (light lines).}
% \label{fig:SpatialMixingKernel}
% \end{figure} 
% \begin{figure}[!h] 
% \centering
% \includegraphics[width=0.5\textwidth]{./Graph/aram2.eps}
% \caption{Spatial mixing kernel; estimated and true kernels are shown by solid and dashed line respectively; top panel is estimation with maximmum scale $j=0$ and bottom panel is estimation with maximum scale $j=1$. }
% \label{fig:SpatialMixingKernelEstimate}
% \end{figure}
% \begin{figure*}[!hb] 
% % \renewcommand*{\thesubfigure}{}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram35.eps}}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram310.eps}}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram350.eps}}\\ 
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram3100.eps}}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram3200.eps}} 
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram3300.eps}}
% \caption{Spatial field at a number of time instants; estimated and true fields are shown by solid and dashed line respectively; the top panel of each plot is the estimate with maximmum scale $j=0$ and the bottom panel is the estimate with maximum scale $j=1$.}
% \label{fig:1DFieldEstimation}
% \end{figure*}

@@

% \begin{figure}[!h] 
% \centering
% \includegraphics[width=0.5\textwidth]{./Graph/aram4.eps}
% \caption{Variance of the field at each spatial location for $j=0$ (dotted line) and $j=1$ (solid line). The observation locations are described with a cross.}
% \label{fig:FieldVariance}
% \end{figure} 

@@

% \begin{figure}[!t] 
% \centering
% \includegraphics[width=0.5\textwidth]{./Graph/aram5.eps}
% \caption{Distribution of the parameter estimate for 2-D Example over 500 runs of the EM algorithm. The dotted line indicates the true parameter.}
% \label{fig:2DHist}
% \end{figure}
% \begin{figure*}[!th] 
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram6a.eps}}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram6b.eps}}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram6c.eps}}\\ 
% \caption{Spatial field at t=5; a: True field; b: Estimated field with maximmum scale $j=0$; c: Estimated field with maximmum scale $j=1$.}
% \label{fig:2DFieldEstimation1}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram7a.eps}}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram7b.eps}} 
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram7c.eps}}\\
% \caption{Spatial field at t=50; a: True field; b: Estimated field with maximmum scale $j=0$; c: Estimated field with maximmum scale $j=1$.}
% \label{fig:2DFieldEstimation2}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram8a.eps}}
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram8b.eps}} 
% \subfigure[][]{\includegraphics[width=2.18in]{./Graph/aram8c.eps}}
% \caption{Spatial field at t=150; a: True field; b: Estimated field with maximmum scale $j=0$; c: Estimated field with maximmum scale $j=1$.}
% \label{fig:2DFieldEstimation3}
% \end{figure*}

@@

\section*{Reducing the computational complexity}\label{App:ReducedRTS}

@@

\begin{equation}\label{eq:Upsilon}

@@

We define $n_x \times n_x$ block matrix $\mathbf U$ such that
\begin{equation}\label{eq:UMatrix}
 [\mathbf U]_{i,j}=\boldsymbol\gamma^{(i,j)\mathsf{T}}
substituting \eqref{eq:UMatrix} in \eqref{eq:Upsilon} we get
\begin{equation}\label{eq:UpsilonU}
 \boldsymbol\Upsilon=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}
and similarly substituting \eqref{eq:UMatrix} in \eqref{eq:Upsilon} we have
\begin{equation}\label{eq:upsilonU}
 \boldsymbol\upsilon=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}
with this substitution the second terms in \eqref{eq:UpsilonU} and \eqref{eq:upsilonU} can be pre-calculated and stored. This lowers the comutational complexity of the estimation algorithm and provides a dramatic increase in estimation speed.
\section*{Cross covariance matrix}\label{App:EnCrossCov}
